Modal Abstractions for Virtualizing Memory Addresses

### 2025-03-21 background.tex
minor bg tweaks, significant adjustment of section 4.2, which was hard to follow because it jumped back and forth between different figures and was missing some explanation

\subsection{Separation Logic}
\label{sec:seplogic}
\todo[inline]{Colin says: I added this b/c prior systems reviewers wanted more verification background...
not sure how I feel about it}
Separation logic~\cite{reynolds02} is a descendant of classic Hoare logic~\cite{hoare69},
where in addition to pre- and postcondition assertions, assertions themselves pick up a
\emph{separating conjunction} operator $\ast$, such that an assertion $A\ast B$ means $A$ and $B$ are true
of disjoint pieces of state. This allows for local reasoning about updates, because it articulates
that updates to the state backing $A$ do not affect the truth of $B$. This is classically demonstrated
through the \emph{points-to} assertion: $l\mapsto v$ asserts that the memory cell at address $l$ holds value $v$:
knowing $x\mapsto 3\ast y\mapsto 4$ and writing through $x$ means information about $y$ is preserved:
$\{x\mapsto 3\ast P\}\;x\mathrel{:=}5\;\{x\mapsto 5 \ast P\}$ can be derived for any $P$.

@@

We suppress many details of \iris assertions for brevity, but briefly note
that because \iris is an embedded framework in Rocq, proofs in \iris-derived
logics (including ours) often encapsulate raw Rocq assertions: $\ulcorner P \urcorner$ is an embedding
of the Rocq assertion $P$ into an \iris assertion (used for things like equality and other
general pure logical assertions). Newly-defined \iris assertions are in fact
Rocq terms of a particular type, rather than being drawn from a fixed vocabulary.
We introduce other details as they arise.

Of particular interest to us, \iris's machinery is largely agnostic to the particular choice of
\emph{resource algebra}\footnote{A modern descendant of the venerable partial commutative monoid used
to abstractly model earlier separation logics~\cite{calcagno2007local}, 
with additions to support the step-indexing required to solve the recursive domain equations
that arise with higher-order ghost state and impredicativity~\cite{birkedal2011step,hobor2010theory}.
}
used to give algebraic semantics to the connectives of the assertion language.
This means we can drop in an alternative model that naturally supports working with respect to
a locally-fixed context, like an address space.

>>>>

\subsection{Separation Logic}
\label{sec:seplogic}
\todo{Colin says: I added this b/c prior systems reviewers wanted more verification background...
not sure how I feel about it}
Separation logic~\cite{reynolds02} is a descendant of classic Hoare logic~\cite{hoare69},
where in addition to pre- and postcondition assertions, assertions themselves pick up a
\emph{separating conjunction} operator $\ast$, such that an assertion $A\ast B$ means $A$ and $B$ are true
of disjoint pieces of state. This allows for local reasoning about updates, because it articulates
that updates to the state backing $A$ do not affect the truth of $B$. This is classically demonstrated
through the \emph{points-to} assertion: $l\mapsto v$ asserts that the memory cell at address $l$ holds value $v$:
knowing $x\mapsto 3\ast y\mapsto 4$ and writing through $x$ means information about $y$ is preserved:
$\{x\mapsto 3\ast P\}\;x\mathrel{:=}5\;\{x\mapsto 5 \ast P\}$ can be derived for any $P$.

@@

We suppress many details of \iris assertions for brevity, but briefly note
a few recurring details used heavily in \iris but not necessarily in traditional
separation logics.
Because \iris is an embedded framework in \rocq, proofs in \iris-derived
logics (including ours) often encapsulate raw \rocq assertions: $\ulcorner P \urcorner$ is an embedding
of the \rocq assertion $P$ into an \iris assertion (used for things like equality and other
general pure logical assertions). Newly-defined \iris assertions are in fact
\rocq terms of a particular type, rather than being drawn from a fixed vocabulary.
We introduce other details as they arise.

\iris includes two forms of implication. The magic wand operator $\wand$ is an affine implication:
$A\wand B$ describes a resource which, if combined with a resource satisfying $A$, will satisfy $B$.
Notably, this implication involves no changes to ghost state. \iris, building on the Views framework~\cite{Dinsdale-Young2013Views},
also includes a \emph{view shift} operator $\vs$ which models updates to ghost state: $A \vs B$
means resources satisfying $A$ may be transformed into resources satisfying $B$, intuitively by updating only ghost state.\footnote{\iris
experts may note that this is \emph{technically} mildly misleading given how \iris's update modalities and weakest precondition
are defined, but
this is adequate intuition for non-experts in \iris to follow their use.
}
\looseness=-1

% Of particular interest to us, \iris's machinery is largely agnostic to the particular choice of
% \emph{resource algebra}\footnote{A modern descendant of the venerable partial commutative monoid used
% to abstractly model earlier separation logics~\cite{calcagno2007local}, 
% with additions to support the step-indexing required to solve the recursive domain equations
% that arise with higher-order ghost state and impredicativity~\cite{birkedal2011step,hobor2010theory}.
% }
% used to give algebraic semantics to the connectives of the assertion language.
% This means we can drop in an alternative model that naturally supports working with respect to
% a locally-fixed context, like an address space.

### 2025-03-21 programlogic.tex
minor bg tweaks, significant adjustment of section 4.2, which was hard to follow because it jumped back and forth between different figures and was missing some explanation

% The idea that the truth of an assertion is relative to an address space has far-reaching consequences.
To support making assertions depend on a choice of address space, we work entirely in a pointwise lifting of Iris's base BI logic,
essentially working with separation logic assertions indexed by a choice of page table root as a $\mathcal{W}_{64}$, which we call $\textsf{vProp }\Sigma$:\footnote{
  \iris experts may notice our \lstinline|-b>| resembles another pointwise lifting already in  \iris~\cite{dang2019rustbelt,dang2022compass}. 
  This similarity is real, but the existing lifting does not work with indexed \coq types
  like our \lstinline|word n| as a domain.
}
\lstinline[language=Coq]|Definition vProp  $\Sigma$ : bi := word 64 -b> iPropI  $\Sigma$|.
This is the (\textsc{Coq}) type of assertions in our logic.
Most constructs in Iris's base logic are defined with respect to any BI-algebra (of \textsc{Coq} type \lstinline|bi|), so automatically
carry over to our derived logic.
However, we must still build up from existing Iris primitives to provide new primitives that depend on the address space --- primarily the notion
of virtual points-to.

@@

  We separate the physical page-table-walk  
  from the virtual pointsto relation, replacing it with ghost state that merely guarantees the address translation would succeed.
  \iris includes a \emph{ghost map} construction which we use to track mappings from
  virtual addresses to the physical addresses they translate to.
  The implementation details in terms of named invariants and various resource algebras are not important to our presentation, so
  we treat the maps abstractly.
   The map includes for each key in the map (i.e.,
  each virtual address) a token $\ghostmaptoken{\gamma}{k}{v}$ whose ownership is required to update that key-value pair in the ghost map
  named $\gamma$. The existence of such a token implies that the actual map $\theta$ tracked by a corresponding $\mathsf{GhostMap}(\gamma,\theta)$
  resource indeed maps $k$ to $v$. These properties are captured by \iris rules using the token to update or draw conclusions about the logical map:\footnote{Iris ghost maps lack established notation in the literature,
  but the syntax we use captures the details of \texttt{iris.base\_logic.lib.ghost\_map}.}
  \begin{mathpar}
  \inferrule{ }{\mathsf{GhostMap}(\gamma,\theta) \ast \ghostmaptoken{\gamma}{\mathsf{pa}}{\mathsf{va}} \vs 
  \mathsf{GhostMap}(\gamma,\theta[\mathsf{pa}\mapsto\mathsf{va'}]) \ast \ghostmaptoken{\gamma}{\mathsf{pa}}{\mathsf{va}'} 
  }
  \and
  \inferrule{ }{\mathsf{GhostMap}(\gamma,\theta) \ast \ghostmaptoken{\gamma}{\mathsf{pa}}{\mathsf{va}} \wand \ulcorner \theta(\mathsf{pa})=\mathsf{Some}(\mathsf{va}) \urcorner }
  \end{mathpar}

  % Updating the virtual-to-physical (ghost) mapping for a virtual address $va$
  % requires full ownership of the token. 
  Then the \emph{virtual memory manager's invariant} ensures that for each $\ghostmaptoken{\gamma}{va}{pa}$ mapping
  in this map, there are \emph{physical} resources sufficient to ensure that address translation for $va$
  will resolve in the hardware to $pa$ via --- $\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$.
  This allows modification of the page tables themselves as long as the changes do not invalidate overall virtual-to-physical
  translation.
  Consistency follows from replacing the page table walk with ownership of the ghost map token for the relevant ghost map
  entry, preventing the overall translation from being changed separately from the virtual points-to.
  \looseness=-1

@@

  % In Iris this is realized by using an authoritative resource algebra: there is a single \emph{authoritative} global copy of the (ghost)
  % map caching virtual-to-physical address translations, and for each entry a read only \emph{partial} ownership of that key-value pair.
  % The resource algebra itself is instantiated as:
  % \[\mathcal{A}\textsf{bsPTableWalk} \stackrel{\triangle}{=} \textsc{Auth} (\; \mathcal{W}_{64} \;\rightarrow_{\textrm{fin}} \;  ( (\textsc{Frac }, \mathord{+}) \times (\textsc{Agree } \Loc,\mathord{=}) ))\]
For clarity we refer to the ghost map summarizing virtual-to-physical translations by 
\centerline{$\mathcal{A}\textsf{bsPTableWalk}(\delta,\theta) \stackrel{\triangle}{=} \mathsf{GhostMap}(\delta,\theta)$}\\
(omitting $\delta$ for brevity when only one is in scope)
and keep this in a per-address-space invariant described shortly.
We then replace the physical
 $\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$ in the virtual points-to definition
with ownership of the token \ghostmaptoken{\delta}{\mathsf{va}}{\mathsf{paddr}}, %ghost-map ($ \sumwalkabs\vaddr\qfrac\paddr$),
guaranteeing that ghost map contains an mapping from the virtual address ($\vaddr$) to a physical address ($\paddr$),
and thus that the per-address-space invariant contains the physical resources that guarantee the hardware resolves the translation.

We place the authorative ownership of the ghost translation $\mathcal{A}\textsf{PTableWalk}$ in a per-address-space invariant
$\mathcal{I}$\textsf{ASpace} (Figure \ref{fig:peraspaceinvariant}). 
$\mathcal{I}$\textsf{ASpace} alone allows changes to the page tables 
that preserve overall virtual-to-physical translations.
When combined with the
fragment stored in the virtual points-to (Figure \ref{fig:virtualpointstosharing}),
changes in the virtual-to-physical translations become possible.

  \begin{figure*}
  \footnotesize
\centerline{$
\begin{array}{l}
  \mathcal{I}\textsf{ASpace}(\ptablestore,m)\stackrel{\triangle}{=} \textsf{ASpace\_Lookup}(\ptablestore,m) \ast \\
 \qquad\qquad\qquad \bigast{(\vaddr, \textsf{paddr})\in \ptablestore}{\exists\;(\textsf{l4e l3e l2e, l1e, paddr})\ldotp \textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\vaddr\textsf{, l4e, l3e, l2e, l1e, paddr})} \\
  \textsf{ where } 
   \textsf{ASpace\_Lookup} (\ptablestore,m) \stackrel{\triangle}{=} \lambda\textsf{ cr3val} \ldotp \; \exists \gammaPred \; \ldotp \ulcorner m \; !!\; \textsf{cr3val} = \textsf{Some } \gammaPred \urcorner \ast
    %\ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}}
    \ptableabswalk{\delta,\theta}
\end{array}
$}
\vspace{-1em}
\caption{Global Address-Space Invariant with a fixed global map of address-space names $m$}
  \label{fig:peraspaceinvariant}
\vspace{-1em}
  \end{figure*}

Our logic is the first to address the existence of multiple address spaces, so also requires
a way to look up
the ghost map of virtual-to-physical translations for an address space (the ghost name $\delta$ in Figure \ref{fig:virtualpointstosharing}).
% for each address space must be locateable from a representation of the specific address space:
% This means we also require a way to locate the map for the current address space from 
We support lookup by the current root page table pointer stored in \lstinline|cr3|.
\looseness=-1

Figure \ref{fig:peraspaceinvariant} describes the resources for each address space.
$\theta$ is the logical map from virtual addresses to the physical addresses they should translate to, 
for the \emph{current} address space, corresponding to the currently installed page tables (the ones
indicated by \lstinline|cr3|). In Figure \ref{fig:virtualpointstosharing}, the existentially-quantified $\delta$
is the ghost name for this map (again, for the current address space).
$m$ is the (full) logical map from the various page table roots to ghost names for the 
per-address-space mappings like $\theta$.
It is tracked by a ghost map named by the ghost name $\delta{}s$ in
Figure \ref{fig:virtualpointstosharing}; there is only one such map in the system.
\looseness=-1
% This is why Figure \ref{fig:virtualpointstosharing} asserts there is some ghost name in that (global) map
% for the current address space.
% The global map itself is tracked by a resource algebra 
% \[\mathcal{A}\textsf{spaces} \stackrel{\triangle}{=} \textsc{Auth} (\; \mathcal{W}_{64} \;\rightarrow_{\textrm{fin}} \;  ( (\textsc{Frac }, \mathord{+}) \times (\textsc{Agree } \textsc{GhostName},\mathord{=}) )) \]

$\mathcal{I}\textsf{ASpace}(\Theta,m)$ is then the resources for the current address space,
ensuring that the current address space is valid (has an entry in $m$ whose corresponding
authoritative map matches $\Theta$), and asserting ownership of the appropriate fractional ownership
of the physical page table walk for each virtual-to-physical translation in $\Theta$ (for the current address space).

>>>>

% The idea that the truth of an assertion is relative to an address space has far-reaching consequences.
To support making assertions depend on a choice of address space, we work entirely in a pointwise lifting of Iris's base BI logic,
essentially working with separation logic assertions indexed by a choice of page table root as a $\mathcal{W}_{64}$, which we call $\textsf{vProp }\Sigma$:\footnote{
  \iris experts may notice our \lstinline|-b>| resembles another pointwise lifting already in  \iris~\cite{dang2019rustbelt,dang2022compass}. 
  This similarity is real, but the existing lifting does not appear to work with indexed \coq types
  like our \lstinline|word n| as a domain.
}
\lstinline[language=Coq]|Definition vProp  $\Sigma$ : bi := word 64 -b> iPropI  $\Sigma$|.
This is the (\textsc{Coq}) type of assertions in our logic.
Most constructs in Iris's base logic are defined with respect to any BI-algebra (of \coq type \lstinline|bi|), so automatically
carry over to our derived logic.
However, we must still build up from existing Iris primitives to provide new primitives that depend on the address space --- primarily the notion
of virtual points-to.

@@

  \begin{figure}
\footnotesize
  \begin{mathpar}
  \inferrule*[right=\footnotesize GhostMapUpdate]{ }{\mathsf{GhostMap}(\gamma,\theta) \ast \ghostmaptoken{\gamma}{\mathsf{pa}}{\mathsf{va}} \vs 
  \mathsf{GhostMap}(\gamma,\theta[\mathsf{va}\mapsto\mathsf{pa'}]) \ast \ghostmaptoken{\gamma}{\mathsf{va}}{\mathsf{pa}'} 
  }
  \and
  \inferrule*[right=\footnotesize GhostMapLookup]{ }{\mathsf{GhostMap}(\gamma,\theta) \ast \ghostmaptoken{\gamma}{\mathsf{va}}{\mathsf{pa}} \wand \ulcorner \theta(\mathsf{va})=\mathsf{Some}(\mathsf{pa}) \urcorner }
  \end{mathpar}
  \caption{\iris rules for ghost maps}
  \label{fig:ghostmaps}
  \end{figure}

  We separate the physical page-table-walk  
  from the virtual pointsto relation, replacing it with ghost state that merely guarantees the address translation would succeed.
  \iris includes a \emph{ghost map} construction which we use to track mappings from
  virtual addresses to the physical addresses they translate to as a piece of ghost state.
  The map includes for each key in the map (i.e.,
  each virtual address) a token $\ghostmaptoken{\gamma}{k}{v}$ whose ownership is required to update that key-value pair in the ghost map
  named $\gamma$. The existence of such a token implies that the actual map $\theta$ tracked by a corresponding $\mathsf{GhostMap}(\gamma,\theta)$
  resource indeed maps $k$ to $v$. These properties are captured by key \iris rules in Figure \ref{fig:ghostmaps}.\footnote{\iris ghost maps lack established notation in the literature,
  but the syntax we use captures the details of \texttt{iris.base\_logic.lib.ghost\_map}.}
  There are other rules, but these two are most important for explaining ghost maps.
  \textsc{GhostMapUpdate} says that ownership of the actual ghost map with ghost name $\gamma$ and map contents $\theta$,
  and a full (non-fractional) token witnessing that $\theta$ maps \textsf{pa} to \textsf{va} permits an update to the ghost map's state,
  changing the map and replacing the token to represent the new value.
  \textsc{GhostMapLookup} allows using the same information to simply conclude that the mapping indicated by the token is true.

  % Updating the virtual-to-physical (ghost) mapping for a virtual address $va$
  % requires full ownership of the token. 
  The \emph{virtual memory manager's invariant} ensures that for each $\ghostmaptoken{\gamma}{\vaddr}{\paddr}$ mapping
  in this map, there are \emph{physical} resources sufficient to ensure that address translation for $\vaddr$
  will resolve in the hardware to $\paddr$ --- via $\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$.
  % This allows modification of the page tables themselves as long as the changes do not invalidate overall virtual-to-physical
  % translation.
  % Consistency follows from replacing the explicit page table walk of Figure \ref{fig:strongvirtualpointsto}'s
  % definition with ownership of the ghost map token for the relevant ghost map
  % entry, as in Figure \ref{fig:virtualpointstosharing}.
  % This allows the overall translation to be updated separately from the virtual points-to.
  \looseness=-1

@@

  % In Iris this is realized by using an authoritative resource algebra: there is a single \emph{authoritative} global copy of the (ghost)
  % map caching virtual-to-physical address translations, and for each entry a read only \emph{partial} ownership of that key-value pair.
  % The resource algebra itself is instantiated as:
  % \[\mathcal{A}\textsf{bsPTableWalk} \stackrel{\triangle}{=} \textsc{Auth} (\; \mathcal{W}_{64} \;\rightarrow_{\textrm{fin}} \;  ( (\textsc{Frac }, \mathord{+}) \times (\textsc{Agree } \Loc,\mathord{=}) ))\]
For clarity we refer to the specific ghost map summarizing virtual-to-physical translations by \\
\centerline{$\mathcal{A}\textsf{bsPTableWalk}(\delta,\theta) \stackrel{\triangle}{=} \mathsf{GhostMap}(\delta,\theta)$}\\
(omitting $\delta$ for brevity when only one is in scope)
and keep this in a per-address-space invariant described shortly.
We then replace the physical traversal
 $\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$ in Figure \ref{fig:strongvirtualpointsto}'s virtual points-to definition
with ownership of the token \ghostmaptoken{\delta}{\vaddr}{\paddr}, %ghost-map ($ \sumwalkabs\vaddr\qfrac\paddr$),
yielding Figure \ref{fig:virtualpointstosharing}'s definition.
This new definition guarantees that the ghost map contains an mapping from the virtual address ($\vaddr$) to a physical address ($\paddr$),
and thus that the per-address-space invariant will contain the physical resources that guarantee the hardware resolves the translation.

  \begin{figure*}
  \footnotesize
\centerline{$
\begin{array}{l}
  \mathcal{I}\textsf{ASpace}(\ptablestore,m)\stackrel{\triangle}{=} \textsf{ASpace\_Lookup}(\ptablestore,m) \ast \\
 \qquad\qquad\qquad \bigast{(\vaddr, \textsf{paddr})\in \ptablestore}{\exists\;(\textsf{l4e l3e l2e, l1e, paddr})\ldotp \textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\vaddr\textsf{, l4e, l3e, l2e, l1e, paddr})} \\
  \textsf{ where } 
   \textsf{ASpace\_Lookup} (\ptablestore,m) \stackrel{\triangle}{=} \lambda\textsf{ cr3val} \ldotp \; \exists \gammaPred \; \ldotp \ulcorner m \; !!\; \textsf{cr3val} = \textsf{Some } \gammaPred \urcorner \ast
    %\ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}}
    \ptableabswalk{\delta,\theta}
\end{array}
$}
\vspace{-1em}
\caption{Per-address-space invariant with a fixed global map of address-space names $m$}
  \label{fig:peraspaceinvariant}
\vspace{-1em}
  \end{figure*}

We place the authorative ownership of the ghost translation $\mathcal{A}\textsf{PTableWalk}$ in a per-address-space invariant
$\mathcal{I}$\textsf{ASpace} (Figure \ref{fig:peraspaceinvariant}). 
$\mathcal{I}$\textsf{ASpace} alone allows changes to the page tables 
that preserve overall virtual-to-physical translations.
When combined with the
fragment stored in the virtual points-to (Figure \ref{fig:virtualpointstosharing}),
changes in the virtual-to-physical translations become possible without needing to track all virtual points-to assertions.

% Figure \ref{fig:peraspaceinvariant} describes the resources for each address space.
In Figure \ref{fig:peraspaceinvariant},
$\theta$ is the logical map from virtual addresses to the physical addresses they should translate to, 
for the \emph{current} address space, corresponding to the currently installed page tables (the ones
indicated by \lstinline|cr3|). In Figure \ref{fig:virtualpointstosharing}, the existentially-quantified $\delta$
is the ghost name for this map (again, for the current address space).
Because our logic is the first to address the existence of multiple address spaces,
it must ensure that $\theta$ actually corresponds to the intended ghost map for the current address space.
This requires looking up the ghost name (i.e., $\delta$) for the current address space,
which is handled by $m$ --- another logical map (tied to ghost state shortly)
from various page table roots to ghost names for the per-address-space ghost maps discussed above.
$\mathcal{I}\textsf{ASpace}(\theta,m)$ then stores the physical resources for the current address space mappings,
ensuring that the current address space is valid (has an entry in $m$ whose corresponding
authoritative map matches $\theta$), and asserting ownership of the appropriate fractional ownership
of the physical page table walk for each virtual-to-physical translation in $\Theta$ (for the current address space).

Virtual points-to assertions must also ensure that they are tied to a valid address space, and because they
use ghost map tokens to ensure the virtual-to-physical translation exists in a specific ghost map, they must also
name that map. In Figure \ref{fig:virtualpointstosharing}, this is done with a ghost map token for \emph{another} ghost map,
one shared across the kernel, holding $m$. Our virtual points-to assertion existentially quanfies the address-space-specific
ghost map $\delta$ mapping virtual addresses to physical addresses, and ensures it exists with a token for the distinguished
global (unique) ghost map named by $\delta{s}$.
% the ghost map of virtual-to-physical translations for a particular address space (the ghost name $\delta$ in Figure \ref{fig:virtualpointstosharing}).
% % for each address space must be locateable from a representation of the specific address space:
% % This means we also require a way to locate the map for the current address space from 
% We support lookup by the current root page table pointer stored in \lstinline|cr3|.
\looseness=-1
